Nonlinear mixed effects models for time series forecasting of smart meter demand

### 2020-12-11 paper.Rmd
Change figures to postscript
- Set fig_crop to false to avoid rasterized images.
- Set dev to postscript.
- Remove semi-transparent elements from plots since this is not
  supported in .eps files.

---
title: "Nonlinear mixed effects models for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Ben Taieb
  othernames: Souhaib
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure electricity demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, there are additional applications that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and nonlinear mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.lazy = FALSE,
  dev = "pdf",
  fig.height = 5,
  fig.width = 8.5
)
library(tidyverse)
library(tsibble)
library(furrr)
library(lubridate)
library(stringr)
library(lme4)
# library(nlme)
library(splines)
library(forecast)
library(leaps)
source("R/load-data.R")
source("R/feature-selection.R")
source("R/test-models.R")
source("R/models-lme4.R")
source("R/plot-helpers.R")
```

@@

```{r temp-wh-scatter, fig.cap="(ref:temp-wh-scatter)"}
main_df %>%
  filter(
    bid %in% c("BID0105", "BID0127"),
    month(ts) %in% c(12, 1, 2),
    period %in% c(48, 96)
  ) %>%
  mutate(time = seconds_to_period(period*15*60),
         time = sprintf("%02d:%02d", time$hour, time$minute)) %>%
  ggplot(aes(x = temperature, y = log(wh), colour = time)) +
  geom_point() +
  stat_smooth(geom='line', alpha=1, se=FALSE, method = "lm",
              formula = y ~ ns(x, df = 3)) +
  labs(
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)",
    colour = "Time of day"
  ) +
  facet_wrap(~bid) +
  theme(legend.position = "bottom") +
  scale_colour_manual(values=ba_palette[c(3,1)])
```

@@

ggplot(mapping = aes(x = temperature)) +
  geom_point(data = p_data, aes(y = log(wh)), shape = "O", alpha = 0.5) +
  geom_line(
    data = p_data %>%
      select(
        temperature, `Linear spline`, `Cubic spline`,
        `Natural spline`
      ) %>%
      gather(var, val, -temperature) %>%
      mutate(var = factor(var, levels = c("Linear spline", "Cubic spline", "Natural spline"))),
    aes(y = val), colour = ba_palette[1]
  ) +
  facet_wrap(~var) +
  labs(
    colour = "Method",
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)"
  ) +
  theme(legend.position = "none")
```

@@

```{r fs-model-selection, fig.cap = "(ref:fs-model-selection)"}
fs_df %>%
  select(month, period, features) %>%
  unnest(cols = c(features)) %>%
  select(month, period, n_vars, loocv) %>%
  mutate(month = month(month, label = TRUE)) %>%
  ggplot(aes(y = loocv, x = n_vars, colour = period, group = period)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~month) +
  scale_colour_gradient2(low = ba_palette[1], mid = ba_palette[4],
                         high = ba_palette[2], midpoint = 48) +
  labs(
    x = "Number of variables",
    y = "LOOCV scores",
    colour = "Period"
  ) +
  scale_x_continuous(breaks=c(0,3,6,9,12)) +
  scale_y_log10()
```

@@

```{r fs-alpha, fig.height = 6, fig.cap="(ref:fs-alpha)"}
fs_df %>%
  select(period, month, vars) %>%
  unnest(cols = c(vars)) %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(selected = tidyr::replace_na(selected, FALSE)) %>%
  group_by(period, vars) %>%
  summarise(n_selected = sum(selected)) %>%
  ungroup() %>%
  filter(vars != "scaled_temperature") %>%
  mutate(selected = if_else(n_selected > 0, TRUE, FALSE)) %>%
  inner_join(predictor_vars) %>%
  mutate(description = factor(description, levels = predictor_vars$description,
                              ordered = TRUE),
         description = fct_rev(description)) %>%
  filter(n_selected > 0) %>% # don't want any fill for 0
  ggplot(aes(x = period, y = description, alpha = n_selected)) +
  geom_tile(fill = ba_palette[1]) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = "Period",
    y = NULL,
    fill = "Predictor type",
    alpha = "Selections"
  ) +
  facet_grid(predictor_type ~ ., space = "free_y", scales = "free_y") +
  scale_x_continuous(breaks = seq(8, 96, 8))
```

>>>>

---
title: "Nonlinear mixed effects models for time series forecasting of smart meter demand"
author:
- familyname: Roach
  othernames: Cameron
  address: Monash University
  email: cameron.roach@monash.edu
  correspondingauthor: true
- familyname: Hyndman
  othernames: Rob
  address: Monash University
- familyname: Ben Taieb
  othernames: Souhaib
  address: University of Mons
abstract: Buildings are typically equipped with smart meters to measure electricity demand at regular intervals. Smart meter data for a single building have many uses, such as forecasting and assessing overall building performance. However, when data are available from multiple buildings, there are additional applications that are rarely explored. For instance, we can explore how different building characteristics influence energy demand. If each building is treated as a random effect and building characteristics are handled as fixed effects, a mixed effects model can be used to estimate how characteristics affect energy usage. In this paper we demonstrate that producing one day ahead demand predictions for 123 commercial office buildings using mixed models can improve forecasting accuracy. We experiment with random intercept, random intercept and slope, and nonlinear mixed models. The predictive performance of the mixed effects models are tested against naive, linear and nonlinear benchmark models fitted to each building separately. This research justifies using mixed models to improve forecasting accuracy and to quantify changes in energy consumption under different building configuration scenarios.
keywords: "time series forecasting, mixed-effects models, smart meters, energy, electricity"
wpnumber: 1/19
jelcodes: C10,C14,C52
blind: false
cover: false
toc: false
bibliography: ["library.bib", "packages.bib"]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    fig_crop: false
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.lazy = FALSE,
  dev = "postscript",
  fig.height = 5,
  fig.width = 8.5
)
library(tidyverse)
library(tsibble)
library(furrr)
library(lubridate)
library(stringr)
library(lme4)
# library(nlme)
library(splines)
library(forecast)
library(leaps)
source("R/load-data.R")
source("R/feature-selection.R")
source("R/test-models.R")
source("R/models-lme4.R")
source("R/plot-helpers.R")
```

@@

```{r temp-wh-scatter, fig.cap="(ref:temp-wh-scatter)"}
main_df %>%
  filter(
    bid %in% c("BID0105", "BID0127"),
    month(ts) %in% c(12, 1, 2),
    period %in% c(48, 96)
  ) %>%
  mutate(time = seconds_to_period(period*15*60),
         time = sprintf("%02d:%02d", time$hour, time$minute)) %>%
  ggplot(aes(x = temperature, y = log(wh), colour = time)) +
  geom_point(shape = "O") +
  stat_smooth(geom='line', se=FALSE, method = "lm",
              formula = y ~ ns(x, df = 3)) +
  labs(
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)",
    colour = "Time of day"
  ) +
  facet_wrap(~bid) +
  theme(legend.position = "bottom") +
  scale_colour_manual(values=ba_palette[c(3,1)])
```

@@

ggplot(mapping = aes(x = temperature)) +
  geom_point(data = p_data, aes(y = log(wh)), shape = "O") +
  geom_line(
    data = p_data %>%
      select(
        temperature, `Linear spline`, `Cubic spline`,
        `Natural spline`
      ) %>%
      gather(var, val, -temperature) %>%
      mutate(var = factor(var, levels = c("Linear spline", "Cubic spline", "Natural spline"))),
    aes(y = val), colour = ba_palette[1]
  ) +
  facet_wrap(~var) +
  labs(
    colour = "Method",
    y = expression(Log ~ normalised ~ electricity ~ (log(Wh / m^2))),
    x = "Temperature (°C)"
  ) +
  theme(legend.position = "none")
```

@@

```{r fs-model-selection, fig.cap = "(ref:fs-model-selection)"}
fs_df %>%
  select(month, period, features) %>%
  unnest(cols = c(features)) %>%
  select(month, period, n_vars, loocv) %>%
  mutate(month = month(month, label = TRUE)) %>%
  ggplot(aes(y = loocv, x = n_vars, colour = period, group = period)) +
  geom_line() +
  facet_wrap(~month) +
  scale_colour_gradient2(low = ba_palette[1], mid = ba_palette[4],
                         high = ba_palette[2], midpoint = 48) +
  labs(
    x = "Number of variables",
    y = "LOOCV scores",
    colour = "Period"
  ) +
  scale_x_continuous(breaks=c(0,3,6,9,12)) +
  scale_y_log10()
```

@@

```{r fs-alpha, fig.height = 6, fig.cap="(ref:fs-alpha)"}
fs_df %>%
  select(period, month, vars) %>%
  unnest(cols = c(vars)) %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(selected = tidyr::replace_na(selected, FALSE)) %>%
  group_by(period, vars) %>%
  summarise(n_selected = sum(selected)) %>%
  ungroup() %>%
  filter(vars != "scaled_temperature") %>%
  mutate(selected = if_else(n_selected > 0, TRUE, FALSE)) %>%
  inner_join(predictor_vars) %>%
  mutate(description = factor(description, levels = predictor_vars$description,
                              ordered = TRUE),
         description = fct_rev(description)) %>%
  filter(n_selected > 0) %>% # don't want any fill for 0
  ggplot(aes(x = period, y = description, fill = n_selected)) +
  geom_tile() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = "Period",
    y = NULL,
    fill = "Selections"
  ) +
  facet_grid(predictor_type ~ ., space = "free_y", scales = "free_y") +
  scale_x_continuous(breaks = seq(8, 96, 8)) +
  scale_fill_gradient2(low = "white", high = ba_palette[1])
```
